---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biostoreCapacity

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/biostoreCapacity)](https://CRAN.R-project.org/package=biostoreCapacity)
[![R-CMD-check](https://github.com/mshilts1/biostoreCapacity/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mshilts1/biostoreCapacity/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `biostoreCapacity` is to attempt to predict when VUMC's institutional resource [BioStore II freezer](https://www.vumc.org/oor/index.php/vumc-biospecimen-storage) will be full and unable to store any additional ECHO biospecimens.    

To estimate this, the data in the model should include:    

* Historical data of the rate of freezer filling from ECHO.   
* Future biospecimen kit builds.
* 
* Expected participant enrollment numbers from ?????? (I DO NOT HAVE THIS YET).   

## Installation

You can install the development version of biostoreCapacity from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("mshilts1/biostoreCapacity")
```

Eventually, I'm going to *attempt* to put this on Shiny so it's easy for anyone to use, but I will still keep the source code transparent on GitHub.   

## What is the total BioStore II capacity?

The simplest equation for calculating BioStore capacity is:    

<center>$\frac{(196,412 + x)}{788,256} + \frac{(212,692 + y)}{438,840} = 1$</center>


where:    
* $196,412$ is the number of ECHO 1.0 ml tubes already stored (or pending) in the BioStore (as of 2025-09-02).   
* $x$ is the number of 1.0 ml tubes still to be collected for ECHO.    
* $212,692$ is the number of ECHO 1.9 ml tubes already stored (or pending) in the BioStore (as of 2025-09-02).    
* $y$ is the number of 1.9 ml tubes still to be collected for ECHO.     
* $788,256$ is the absolute maximum number of 1.0 ml tubes that can be stored in the BioStore (assuming 0 1.9 ml tubes).     
* $438,840$ is the absolute maximum number of 1.0 ml tubes that can be stored in the BioStore (assuming 0 1.9 ml tubes).

Both $x$ and $y$ can increase, but as one increases the capacity for the other decreases. The total capacity cannot exceed 1, or 100%.

## What should go into the forecast model of when the BioStore will be at capacity?

How can we estimate both $x$ and $y$ above, and when the capacity of the BioStore will be full?    

**This is the data that I think we need to predict when the BioStore will be full:**     

* Historical data (time series data on number of ECHO tubes store in the BioStore). $\color{green}{\text{✓}}$  
* Expected number of kits that will be collected by kit type over time. $\color{red}{\text{✘}}$   
* Expected number of kits over time needs to include ability to handle complexities introduced due to "specialized" kits, which are not collected by all sites. $\color{yellow}{\text{◎}}$
* Number of tubes in current kit builds per each kit type. $\color{green}{\text{✓}}$  
* Proportion of tubes from each kit type expected to be sent back to the biorepository. (e.g., may get only a tiny bit of urine from young babies, and so won't get all three 1.9ml tubes back.). $\color{yellow}{\text{◎}}$  
 
Green checkmark ($\color{green}{\text{✓}}$ ) means we have that data, a yellow dot ($\color{yellow}{\text{◎}}$) means it's speculative estimated data that we can kind of guess at, while a red x ($\color{red}{\text{✘}}$ ) means **WE AS THE LAB CORE** are missing that specific information but it does exist.    

Here's an idea of the kind of formula I'm thinking of, where $FF$ is "Freezer Filling":   

First, we can attempt to make a model using the historical data of ECHO submissions to the BioStore:    
<center>$FF_{t+1} = f(FF_{t} + FF_{t-1} + FF_{t-2} + \cdots + error)$</center>

Second, we know there were changes to the ECHO protocol that will mean the historical rate of data:    
<center>$FF = f($ future enrollment, biospecimen collection schedule, number of tubes in each kit, rate of loss, $error)$</center>


## Usage


```{r example}
library(biostoreCapacity)
```

## Load in data above that we do have

### Historical data

```{r historical_data}
historical_data <- readHistorical()
historical_data_long <- longifyReadHistorical() # same thing as above, but in "long" format for easier plotting
```

Plot rate of accessioning over time:

```{r plot_history}
library(ggplot2)
ggplot(historical_data_long, aes(x = date, y = total, colour = tube_type)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  ylab("Cumulative Tubes Submitted to BioStore") +
  xlab("") +
  scale_x_date(date_breaks = "2 month", date_labels = "%b %y")
```

### Future kit builds and biospecimen collection protocol

```{r biospec_protocold}
biospecimen_collections <- readCollections()
biospecimen_collections
```

There's currently some information in `readCollections()` that's speculative and outright intended to be tunable by an end user.   

**Information in `readCollections()` that can be assumed to be true and constant for the sake of building the model:**      

* All columns with information about the kit builds:    
    * `collection_id`, `kit_type`, `biospecimen_type`, `participant`, `tube_size`, `tubes_per_kit`.   
* All columns about the biospecimen collection timeline:    
    * `visit`, `specimen_type`, `y_2025`, `y_2026`, `y_2027`, `y_2028`, `y_2029`, `y_2030`, `specialized_obesity`, `specialized_chemphys`, `specialized_lifestyle`.    


**Speculative columns all contain the word "proportion" in the name:**     

    * `proportion_from_kit_collected`, `y_2025_proportion`, `y_2026_proportion`, `y_2027_proportion`, `y_2028_proportion`, `y_2029_proportion`, `y_2030_proportion`, `specialized_obesity_proportion`, `specialized_chemphys_proportion`, `specialized_lifestyle_proportion`.   
    

A more thorough description of every column in `readCollections()`:

1. collection_id: a unique ID. a concatenation of kit_type, tube_size, visit, and specimen_type.   
2. kit_type: concatenation of biospecimen_type and participant.       
3. biospecimen_type: type of biospecimen being collected (e.g., urine or blood, etc...).    
4. participant: specimen to be collected from an ECHO child, child's mother, or child's mother's current partner.       
5. tube_size: whether tube is 1.0mL or 1.9mL (1.9mL tubes take up more space in the BioStore.).      
6. tubes_per_kit: number of tubes of specified size in that specific kit.       
7. proportion_from_kit_collected. what proportion of tubes in that kit are we expecting to be returned to be stored in the BioStore? for example, newborn babies may not produce enough urine for all three 1.9mL tubes to be filled and returned. maybe only two will be returned.   
8. visit. time point in child's or child's mother's life when specimen is collected.   
9. visit_logical_order. not really that useful here, but orders the visit column by the logical order of an ECHO child's life (sort of; due to preconception protocol, that gets complicated).   
10. specimen_type. is specimen considered by ECHO a core, preconception, or specialized specimen? this is important because sites are expected to at least try to collect every core specimen, while sites are only allowed to collected specific specialized specimens. the preconception specimens are somewhere in the middle.   
11. y_2025. 1 (yes)/ 0 (no) column. is this specific specimen to be collected in calendar year 2025?    
12. y_2026. same as for y_2025, but calendar year 2026.      
13. y_2027. same as for y_2025, but calendar year 2027.       
14. y_2028. same as for y_2025, but calendar year 2028.       
15. y_2029. same as for y_2025, but calendar year 2029.       
16. y_2030. same as for y_2025, but calendar year 2030.       
17. y_2025_proportion: this is set to 0.25 because there's only about 25% of calendar year 2025 left.
18. y_2026_proportion: leave at 1 unless there's some reason to think specimens won't be collected for all of 2026.             
19. y_2027_proportion: leave at 1 unless there's some reason to think specimens won't be collected for all of 2027.         
20. y_2028_proportion: leave at 1 unless there's some reason to think specimens won't be collected for all of 2028.       
21. y_2029_proportion: leave at 1 unless there's some reason to think specimens won't be collected for all of 2029.       
22. y_2030_proportion: leave at 1 unless there's some reason to think specimens won't be collected for all of 2030.       
23. specialized_obesity: is that biospecimen being collected by sites where the PI selected "obesity" as an outcome of interest?   
24. specialized_obesity_proportion. proportion of sites    
25. specialized_chemphys.   
26. specialized_chemphys_proportion.   
27. specialized_lifestyle.   
28. specialized_lifestyle_proportion.   
29. notes.    
