---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biostoreCapacity

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/biostoreCapacity)](https://CRAN.R-project.org/package=biostoreCapacity)
<!-- badges: end -->

The goal of biostoreCapacity is to ...

## Installation

You can install the development version of biostoreCapacity from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("mshilts1/biostoreCapacity")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(biostoreCapacity)
## basic example code
```


The simplest equation for calculating BioStore capacity is:    

$(a + b)/788256 + (c + d)/438840 = 1$

where:    
$a$ is the current number of 1.0 ml tubes already stored (or pending storage) in the BioStore.   
$b$ is the number of 1.0 ml tubes still to be collected for ECHO.
$c$ is the current number of 1.9 ml tubes already stored (or pending storage) in the BioStore
$d$ is the number of 1.9 ml tubes still to be collected ECHO. 
$788,256$ is the absolute maximum number of 1.0 ml tubes that be stored in the BioStore (assuming 0 1.9 ml tubes). 
$438,840$ is the absolute maximum number of 1.0 ml tubes that be stored in the BioStore (assuming 0 1.9 ml tubes). 
